Epistasis is widespread in the genetic control of transcription in humans

### 2013-05-30 docs/manuscript/nature_letter.tex
awaiting permutations to finish
chromosome interactions completed. need to see which ones are hits

\begin{abstract}
A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms, and some examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits, but we hypothesised that it has previously been too technically difficult to detect due to statistical power and computational issues. Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, evidence for multiple instances of epistasis is found. In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions. We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset. Several genes are controlled by multi-locus epistatic interactions whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs. For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. Additionally, 69 of the genetic interactions overlap with known chromosome interactions ($p < 1.0 \times 10^{-4}$). This study presents the first strong evidence for the widespread existence of epistasis emerging from natural genetic variation in humans.
\end{abstract}

@@

\subsection{Discussion}
Most genetic studies focus on the additive effects of polymorphisms, and there is a

>>>>

\begin{abstract}
A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms, and some examples have been shown in other species, few convincing examples exist for epistasis amongst natural polymorphisms in human traits. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits, but we hypothesised that it has previously been too technically difficult to detect due to statistical power and computational issues. Here we show that, using advanced computation techniques and a gene expression study design where any effects are expected to be large, evidence for multiple instances of epistasis is found. In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing the expression of 249 genes is controlled by 549 significant pairwise epistatic interactions. We attempted replication in two independent datasets and 421 show evidence of significance in at least one dataset. Several genes are controlled by multi-locus epistatic interactions whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs. For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. Additionally, 69 of the genetic interactions overlap with known chromosome interactions ($p < 1.1 \times 10^{-70}$). This study presents the first strong evidence for the widespread existence of epistasis emerging from natural genetic variation in humans.
\end{abstract}

@@

\subsection{Discussion}

There are a likely to be a large number of mechanisms by which epistasis may occur, as evidenced by the functional analaysis of our discovery set. Therefore it is not surprising that epistasis should be widespread. Further research is required to elucidate specific mechanisms by which statistical epistasis manifests from molecular action, and to quantify the contribution of epistasis to the genetic variance of complex traits.
